---
title: "Ride Share Base Code"
author: "David Fan"
date: "5/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

Ride Sharing Economy is an interesting one that deeply tied in our lives today.

My Goal is two fold; On a utility perspective, I would want to build a model that can help people better estimate how much commuting expense they'll use through ride share services. On an exploration perpspective I am also quite interested in seeing what are the most significant factors the model considered when making the price prediction.

# Code Proper

This Document contains the code for the data aggregation and modelling
Data Source:
https://data.cityofchicago.org/Transportation/Transportation-Network-Providers-Trips/m6dm-c72p/data
https://www.wunderground.com/history/daily/us/il/chicago/KMDW/date/2019-5-7?cm_ven=localwx_history
http://www.robparal.com/ChicagoCommunityAreaData.html
https://www2.illinois.gov/cms/personnel/employeeresources/pages/stateholidays.aspx

```{r results='hide', message=FALSE, warning=FALSE}
#Load Library
library(readr)
library(dplyr)
library(chron)
library(stringr)
library(ggplot2)
setwd('C:/Users/fzy20/Downloads')

#Load Datasets
trips = read_csv("Transportation_Network_Providers_-_Trips.csv")
holiday = read_csv("C:/Users/fzy20/Documents/Penn/Chicago Holidays.csv")
areaInfo = read_csv("C:/Users/fzy20/Documents/Penn/CommAreaInfo.csv")
areaCode = read_csv("C:/Users/fzy20/Downloads/CommAreas.csv")
weather = read_csv("C:/Users/fzy20/Documents/Penn/Weather Chicago.csv")

#Clean Data, drop data without proper location reference
trips_cleaned = trips[!is.na(trips$`Pickup Centroid Location`),]
trips_cleaned = trips_cleaned[!is.na(trips_cleaned$`Dropoff Centroid Location`),]


#Sample Data for analysis due to computational constraint
trips_selected = trips_cleaned[sample(1:nrow(trips_cleaned),size = 500000),]

#Pick relevant area information
areaInfo = areaInfo[,c(1:7)]

#Feature Engineer the Date and the Time
trips_selected$Start_Date = sapply(trips_selected$`Trip Start Timestamp`,function(x) return(strsplit(x, " ")[[1]][1]))
trips_selected$Start_Time = sapply(trips_selected$`Trip Start Timestamp`,function(x) return(strsplit(x, " ")[[1]][2]))
trips_selected$Start_AM_PM = sapply(trips_selected$`Trip Start Timestamp`,function(x) return(strsplit(x, " ")[[1]][3]))
trips_selected$End_Date = sapply(trips_selected$`Trip End Timestamp`,function(x) return(strsplit(x, " ")[[1]][1]))
trips_selected$End_Time = sapply(trips_selected$`Trip End Timestamp`,function(x) return(strsplit(x, " ")[[1]][2]))
trips_selected$End_AM_PM = sapply(trips_selected$`Trip End Timestamp`,function(x) return(strsplit(x, " ")[[1]][3]))
trips_selected$Start_Date = as.Date(trips_selected$Start_Date, format = "%m/%d/%Y")
trips_selected$End_Date = as.Date(trips_selected$End_Date, format = "%m/%d/%Y")
head(trips_selected)
```

```{r message=FALSE}
#clean holiday datasets
holiday$`2018` = sapply(holiday$`2018`,function(x) return(paste(strsplit(x, ",")[[1]][2]," 2018")))
holiday$`2019` = sapply(holiday$`2019`,function(x) return(paste(strsplit(x, ",")[[1]][2]," 2019")))

#Change the vairable type to a date variable
holiday$`2018` = as.Date(holiday$`2018`,format = " %B %d %Y")
holiday$`2019` = as.Date(holiday$`2019`,format = " %B %d %Y")

#Concatenate the columns
holiday_1 = holiday[,c(1,2)]
holiday_2 = holiday[,c(1,3)]

#Rename Columns
colnames(holiday_1) = c("holiday","Date")
colnames(holiday_2) = c("holiday","Date")

#Put together the columns
holiday = rbind(holiday_1,holiday_2)
holiday$is_holiday = 1
head(holiday)
```

```{r}
#Reframed weather data
weather$Date = as.Date(weather$Date, format = "%m/%d/%Y")
#Merge Weather and Holiday
weather_holiday = merge(weather, holiday, by.x = "Date",by.y = "Date", all.x = TRUE)
head(weather_holiday)
```

```{r}
#Merge area informations
area_complete = merge(areaCode, areaInfo, by = c("Area Number","Name"))
head(area_complete)
```

```{r}
#Merging Relevant datasets
trips_with_weather_holiday = merge(trips_selected, weather_holiday, by.x = 'Start_Date', by.y = 'Date', all.x = TRUE)
trip_final = merge(trips_with_weather_holiday, area_complete, by.x = "Pickup Community Area", by.y = "Area Number")
trip_final[is.na(trip_final$is_holiday),'is_holiday'] = 0
#Display final datasets
head(trip_final)
```

```{r}
#Unselect columns that will not aid the pre-ride insights and predictive model
trip_final_tuned = trip_final %>% select(-c("Trip Start Timestamp","Trip End Timestamp","Dropoff Centroid Location",
                         "Pickup Centroid Location","Shared Trip Authorized","Trip ID", "Pickup Census Tract", "Dropoff Census Tract"))
#Add additional month and weekday feature
trip_final_tuned$Month = months(trip_final_tuned$Start_Date)
trip_final_tuned$Weekdays = weekdays(trip_final_tuned$Start_Date)
#reformat the median Household Income Variable
trip_final_tuned$Median_Household_Income = sapply(trip_final_tuned$Median_Household_Income, function(x) gsub("\\$|,","",x))
trip_final_tuned$Median_Household_Income = as.integer(trip_final_tuned$Median_Household_Income)
#convert feature to time
trip_final_tuned$Start_Time <- chron(times=trip_final_tuned$Start_Time)
head(trip_final_tuned)
```

```{r}
colnames(trip_final_tuned)
```

```{r}
#Create distance variable
trip_final_tuned$distance = sqrt((trip_final_tuned$`Pickup Centroid Longitude`- trip_final_tuned$`Dropoff Centroid Longitude`)^2 + (trip_final_tuned$`Pickup Centroid Latitude`- trip_final_tuned$`Dropoff Centroid Latitude`)^2)
#remove observations that has no time or fare
trip_final_tuned = trip_final_tuned[!is.na(trip_final_tuned$Fare),]
trip_final_tuned = trip_final_tuned[!is.na(trip_final_tuned$`Trip Seconds`),]

#Reconstruct column name to aid data training
names(trip_final_tuned)<-str_replace_all(names(trip_final_tuned), c(" " = "" , "," = "","-" = "_" ))
trip_final = trip_final_tuned[!is.na(trip_final_tuned$Fare),]
head(trip_final)
```
# Selected Visuals
```{r message = FALSE}
ggplot(data = trip_final, aes(x = ordered(Weekdays, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", 
"Friday", "Saturday", "Sunday")), y = Fare)) + geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r message = FALSE}
ggplot(data = trip_final[trip_final$Start_AM_PM == "AM",], aes(x =factor(Start_Time), y = Fare)) + geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
```{r}
ggplot(data = trip_final[trip_final$Start_AM_PM == "PM",], aes(x =factor(Start_Time), y = Fare)) + geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
ggplot(data = trip_final, aes(x =Start_AM_PM, y = Fare)) + geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
```{r}
#interesting
ggplot(data = trip_final, aes(x =distance, y = Fare)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + geom_smooth(method = "lm")
```
```{r}
ggplot(data = trip_final, aes(x =AvgTemp, y = Fare)) + geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r}
ggplot(data = trip_final[trip_final$Start_AM_PM == "AM",], aes(x = Median_Household_Income, y = Fare)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+ geom_smooth(method = "lm")
```

```{r}
trip_final_tuned[trip_final_tuned$Start_Time < chron(times ="6:00:00"),"Start_AM_PM"] = sapply(trip_final_tuned[trip_final_tuned$Start_Time < chron(times ="6:00:00"),"Start_AM_PM"], function(x) paste(x,"early")) 
trip_final_tuned[trip_final_tuned$Start_Time >= chron(times ="6:00:00"),"Start_AM_PM"] = sapply(trip_final_tuned[trip_final_tuned$Start_Time >= chron(times ="6:00:00"),"Start_AM_PM"], function(x) paste(x,"late")) 
```

```{r}
trip_final_tuned=trip_final_tuned %>% mutate_if(is.character, as.factor)
head(trip_final_tuned)
```

```{r}
#train-test split
index = sample(1:nrow(trip_final_tuned), 50000)
train = trip_final_tuned[-index,]
test = trip_final_tuned[index,]
```

```{r}
#Train LM Model on Fare
prelim_model = lm(Fare~TripsPooled + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays, data = train)
```

```{r}
#Train LM Model on Time
prelim_model_time = lm(TripSeconds~TripsPooled + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays, data = train)
```


```{r}
#View model summary Fare
summary(prelim_model)
```

```{r}
#View Model Summary Time
summary(prelim_model_time)
```

```{r}
#Accuracy Report
score = predict(prelim_model, test)
print(paste("price error: ",sqrt(mean((score - test$Fare)^2))))
time = predict(prelim_model_time, test)
print(paste("price error: ", sqrt(mean((time - test$TripSeconds)^2))))
```

```{r}
library(rpart)
#Train Rpart Model on Fare
tree_model_1 = rpart(Fare~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 1))
tree_model_5 = rpart(Fare~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 5))
tree_model_10 = rpart(Fare~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 10))
```

```{r}
#Train Rpart Model on Time
tree_model_time_1 = rpart(TripSeconds~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 1))
tree_model_time_5 = rpart(TripSeconds~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 5))
tree_model_time_10 = rpart(TripSeconds~TripsPooled + Start_Time + Start_AM_PM + AvgTemp + MaxWindSpeed + MaxHumidity + AvgPrecipation + is_holiday + Population + Median_Household_Income + Pop_Below_Poverty + Owner_Occupied + Renter_Occupied_House + distance + Weekdays + Month, data = train,control =list(maxdepth = 10))
```

```{r}
#accuracy
score = predict(tree_model_1, test)
print(paste("maxdepth = 1, price error: ",sqrt(mean((score - test$Fare)^2))))
score = predict(tree_model_5, test)
print(paste("maxdepth =5, price error: ",sqrt(mean((score - test$Fare)^2))))
score = predict(tree_model_10, test)
print(paste("maxdepth = 10, price error: ",sqrt(mean((score - test$Fare)^2))))
time = predict(tree_model_time_1, test)
print(paste("maxdepth = 1, time error: ",sqrt(mean((time - test$TripSeconds)^2))))
time = predict(tree_model_time_5, test)
print(paste("maxdepth = 5, time error: ",sqrt(mean((time - test$TripSeconds)^2))))
time = predict(tree_model_time_10, test)
print(paste("maxdepth = 10, time error: ",sqrt(mean((time - test$TripSeconds)^2))))
```

```{r}
#Pick better model = LM
save(prelim_model, file = "price_model.rda")
save(prelim_model_time, file = "time_model.rda")
```

```{r}
#generate additional datasets to be used for dashboard
Community_location_df = trip_final_tuned %>% group_by(Name) %>% dplyr::summarize(Area_num = mean(`PickupCommunityArea`),latitude = mean(`PickupCentroidLatitude`),
                                                    longitude = mean(`PickupCentroidLongitude`))
```

